Allee Effect and Cultural Evolution

### 2015-09-15 manuscript/draft.tex
(Semi)-Final Draft

\begin{abstract}
Several social learning strategies guide the selection of cultural variant through an evaluation of the expected benefits associated with each observed alternatives. The assumption of the social learner is that the payoff received by the transmitter, while displaying a given variant, is a proxy for  inferring the qualitative features of the variant itself. Here we explore the theoretical implications when: 1) the payoff attributed to a given cultural variant directly affects the reproductive success of its bearer;  and 2) payoff is density dependent, that is its expected amount is a function of the number of individuals possessing the same variant. We consider situations where such density dependence is both direct and inverse, with minimum and maximum critical thresholds bounding the range of positive payoff and a peak value at midpoint. This non-linear relationship, known as Allee effect, portrays a large number of behavioural traits that are enhanced by cooperation or mutual facilitation, but are  constrained by some form of carrying capacity.  We explore the evolutionary dynamics of a single trait-two variants model, each exhibiting a different Allee effect. We use a modified Lotka-Volterra model of payoff-biased social learning, and show that different degrees of reliance on social learning can exhibit a variety of equilibria, including episodes of reversion, stable co-existence, and fixation.

@@

The last three decades have witnessed a steady growth of formal models of cultural evolution. Early works inspired by population biology \citep{cavallisforza_feldman_1981,boyd1985} have laid the foundation of a now mature trans-disciplinary field which unites the social and biological sciences \citep{mesoudi_etal_2006, mcelreath_and_boyd_2007}. The rich cross-fertilisation between different subfields has promoted the designing of new analytical techniques as well as new uses for existing methods, but more crucially have prompted the development of theoretical models of how culture is acquired and transmitted between individuals~\citep{mesoudi_cultural_2015}. %We need probably a list of references, say about 5-7 papers?

@@

This paper examines a group of social learning strategies generally referred to as \emph{payoff-based transmission} \citep{schlag1998,kendal_etal_2009,lake_and_crema_2012,baldini2013,kandler_and_laland_2013,crema_lake_inpress}. Here the probability of copying a cultural variant $i$ is directly proportional to a payoff $s$ exhibited by individuals possessing $i$. In other words the social learner assumes that the success of an individual (measured in $s$) is a function of the observed behavioural trait, that is $s=f(i)$. Several variants of this learning strategies have been proposed and explored, including: \emph{imitate the best}, where the social learner identifies and copies the trait possessed by the individual with the highest payoff; \emph{compare means}, where the social learner compare the average payoff of different strategies and chooses the highest; and the \emph{proportional imitation rule}, where the probability of adopting a variant is proportional to the difference in payoff between the demonstrator and the learner~\citep{schlag1998,baldini2013,crema_lake_inpress}.

@@

The uncertainty in the payoff signal is derived by the presence of interacting cultural and physical traits, as well as the environment where the trait is manifest. Thus the function $s=f(i)$ should be better described as $s=f(i,\epsilon)$, where $\epsilon$ represent all additional hidden and latent variables that contribute, along with the cultural variant of interest $i$, to the payoff signal $s$. A trivial example is a fisherman who might evaluate the performance of a new hook (the target cultural trait) using as a cue the number of fish captured (the payoff signal). The correlation between the two is partly derived by the efficiency of the hook, but also by the rod, the choice of the bait, the amount of available prey, the choice of the fishing spot, the physical strength of the fisherman, and chance. In many cases we should expect that the social teacher and learner share many of these interacting traits as well as the environment where these are displayed. In other words, cultural and ecological inheritance should in part reduce the impact played by $\epsilon$, and consequently the uncertainty in the payoff-signal. There are however situations where even with most variables hold constant the payoff is subject to high variability. Perhaps one of the most relevant examples are 
are contexts where the success of a variant is strongly correlated with the absolute number of individuals $n_i$ possessing the same trait ($i$) within a population. This density dependence, which can be described by the payoff function $f(i,n_i)$, implies that the adoption (or abandonment) of a trait can alter its expected payoff-signal, potentially causing a cascade of feedback effects.

Several behavioural traits are expected to show such a density dependence. For example, behaviour linked to the exploitation of limited resources will exhibit a decline in payoff when density exceeds a critical threshold (i.e carrying capacity). In this case new optimal variants are likely to spread rapidly when they are rare, but as the number of individuals possessing them approach their respective carrying capacities, their beneficial effects will start to decline, the payoff will decrease, and eventually rarer suboptimal variants might be preferred (see Lake and Crema \citeyear{lake_and_crema_2012} for an extensive analysis).

Density dependence does not, however, necessarily imply only a negative correlation between population size and payoff. In many ecological contexts per capita growth rate is positively correlated with population density, thus resulting into an inverse density dependence. This phenomena, known as \emph{Allee} effect \citep{allee1958,courchamp_etal_1999}, is generally explained by a range of processes including cooperation, mutual facilitation, and environmental conditioning; all processes that we expect to observe  in many cultural traits. The Allee effect has been extensively studied  from  empirical and theoretical standpoints, and its implications assessed for a variety of phenomena, including habitat selection \citep{greene_and_stamps_2001}, resource competition \citep{jang2013},  dispersal \citep{steele_2009}, shifts in settlement pattern \citep{crema_2014}, colonisation events,  \citep{kennet_etal_2006}, genetic diversity \citep{roques_etal_2012} and extinction of languages \citep{sutherland_2003} . Yet, despite this wide array of applications, we are unaware of any theoretical studies exploring how the simultaneous presence of direct and inverse density dependence in the payoff signal can affect the evolutionary dynamics of cultural transmission.

While the concept of Allee effect is well established in population ecology (see Kramer et al \citeyear{kramer_etal_2009} for a systematic review), there is no concept with direct equivalence in the social sciences. That said, there are multiple, established lines of evidence that support  the existence of both direct and inverse density dependence in cultural and behavioural traits. For example several ethnographic studies on small scale societies show how the per-capita return of hunting groups show higher yield when the group size is not too small nor too big~ \citep{hill_and_hawkes_1983,janssen_and_hill_2014}. Other instances of inverse density dependence include population level activities such as the use of anthropogenic fire~\citep{bird2013} or selective hunting~\citep{dods_2002} which often require a population size above a certain critical density in order to promote sufficient niche enhancement and a consequent increase in payoff~\citep{rowley-conwy_and_layton_2011}. In a broader sense any cultural trait that benefits from direct or indirect cooperation, 
promote positive niche construction \citep{vandermeer_2008}, and requires a critical mass \citep{rogers_2003} are likely to exhibit some degree of Allee effect. The implication is thus not limited to traits that directly enhance reproductive fitness, and can include language~\citep{kandler2009}, organisational populations~\citep{caroll_and_hannan_1989} and communication technologies \citep{van_slyke_perceived_2007}.

Here we explore a difference equation model of cultural transmission with traits exhibiting an Allee effect.  We assume that: 1) the social learning strategy is payoff-based; and 2)  the payoff is equivalent to the reproductive fitness of each individual. We demonstrate that direct and inverse density dependence can strongly affect the evolutionary trajectory of the system, leading to changes in the conditions required to reach specific equilibria, as well as to the emergence of unstable patterns characterised by temporary or permanent episodes of shifts between alternative cultural variants.

\section{The Model}
We consider the evolutionary dynamics of two mutually exclusive cultural variants which payoff functions are characterised by an Allee effect. We assume a non-fixed population of size $N$, subdivided into two subpopulations with sizes $a$ and $b$, each associated with the two cultural variants. We further assume that $a$ and $b$ can change as a result of intrinsic population growth/decline as well as shifts from one to the other as a result of cultural transmission. The core model can be depicted with the following coupled difference equation \eqref{eq1}:

@@

We investigate the equilibrium conditions for different initial population sizes of $a$ and $b$ and different settings of $z$, $A$, and $K$, ensuring that in all cases $K_b \geq K_a$. We assumed that two strategies differed only in their critical densities and carrying capacities but not on their intrinsic payoffs. To achieve this we set $r_b=\rho r_a$, where $\rho$ is a constant\footnote{Given two variants with the same baseline growth rate (i.e. $r_a=r_b$), equation \eqref{eq2} returns a lower $R$ for the variant with higher $A$ and $K$. In order to avoid this we increased the baseline of the strategy with lower $R$ by a factor $\rho$, given by the following equation:
\begin{align*}
\rho=\frac{A_bK_b(K_a-A_a)^2}{A_aK_a(A_b-K_b)^2}
\end{align*}
  } 
that ensures that the two variants have always the same $R_{max}$. Given that equations \ref{eq1}-\ref{eq4} have no analytical solution that we are aware of, we identified equilibrium conditions via simulation (source codes of the simulation can be found on the following repository \textcolor{red}{PUT LINK HERE})

@@

\begin{figure*}
  \centering
      \includegraphics[width=\textwidth]{./figures/figure5}
  \caption{Basins of attraction for $z=0.05$ (top) and $z=0.2$ (bottom); $\lambda=1$ (left), $\lambda=0.75$ (middle) and $\lambda=0.5$}
    \label{fig:overlap}
\end{figure*}

@@

Our analysis examined the impact of direct and inverse density dependence in payoff-based social learning strategies. The results showed that the coexistence of both positive and negative feedback mechanism can generate a wide range of patterns where small changes in behaviour and initial conditions can both deeply affect the evolutionary trajectories of the system.

social learning can alter the basins of attraction of stable nodes, leading to different dynamics and equilibria from the very same initial conditions. If there is no reliance on social learning, the critical density separates the ultimate equilibrium of the system: if the initial condition is below this threshold the variant is destined to go extinct, whilst if it is above this value will eventually reach carrying capacity. When social learning is present this boundary shifts, variants can survive even if its initial condition is below critical density and equally can reach extinction even if above this threshold. This scenario is further enhanced when the two variants are not independent but are in direct competition. In this case, social learning can further enhance the advantage of one variant over the other.

When the reliance on social learning is comparatively high, and the two variants have a substantial overlap in their range of positive payoff densities (i.e. they share similar critical densities and carrying capacities), the system does not always reach an stable equilibrium, but can instead exhibit cyclical fluctuations between the two variants. In other words the consequences of social learning is not absorbed by the system, and instead a negative feedback process leads to continuous switch from one variant to the other.

The implications of our model can provide some insights in the evolutionary history of cultural traits exhibiting density dependence. Perhaps the most notable example is the spread of keystone innovations in subsistence strategies which led to the adoption of variants enabling larger carrying capacity, often at the cost of higher critical densities. The spread of behavioural traits such as the know-how of the domestication process~\citep{barker2006} the use of new subsistence tools~\citep{petraglia_population_2009} and processing techniques~\citep{molleson1993}, as well as the inclusion/exclusion and extensification/intensification of different prey species are just some of the most notable examples. In many of these examples the initial spread of the innovation generated a positive feedback process, either through enhanced possibility of direct cooperation, or some form of mutual facilitation often indirectly emerging from positive niche construction. At the same time, ethnographic and archaeological evidence show how many instances of critical transitions where not smooth, often leading to the coexistence of alternative strategies as well as episodes of a return to a previously abandoned cultural traits. Several ethnographic examples showcase small-scale societies alternating between different subsistence modes over a comparatively brief period of time \citep{layton1991,mace_1993,oota_et_al_2005}, while increasing archaeological evidence suggest that the transition from hunting-gathering to farming was not unidirectional, but with instances of temporary reversions \citep{rowley2001,stevens_fuller_2012} %ideally we need a couple of more citations here?
While in most cases these are anecdotal evidence rather than testable empirical cases, the insights of our model suggests that different levels of reliance on social learning could alone portray some of the evolutionary trajectories observed in these systems.

>>>>

\begin{abstract}
Several social learning strategies guide the selection of cultural variant through an evaluation of the expected benefits associated with each observed alternatives. The assumption of the social learner is that the payoff received by the transmitter, while displaying a given variant, is a proxy for  inferring the qualitative features of the variant itself. Here we explore the theoretical implications when: 1) the payoff attributed to a given cultural variant directly affects the reproductive success of its bearer;  and 2) payoff is density dependent, that is its expected amount is a function of the number of individuals possessing the same variant. We consider situations where such density dependence is both direct and inverse, with minimum and maximum critical thresholds bounding the range of positive payoff and a peak value at midpoint. This non-linear relationship, known as Allee effect, portrays a large number of behavioural traits that are enhanced by cooperation or mutual facilitation, but are constrained by finite amount of resources.  We explore the evolutionary dynamics of a single trait-two variants model, each exhibiting a different Allee effect. We use a modified Lotka-Volterra model of payoff-biased social learning, and show that different degrees of reliance on social learning can exhibit a variety of equilibria, including episodes of reversion, stable co-existence, and fixation.

@@

The last three decades have witnessed a steady growth of formal models of cultural evolution. Early works inspired by population biology \citep{cavallisforza_feldman_1981,boyd1985} have laid the foundation of a now mature trans-disciplinary field which unites the social and biological sciences \citep{mesoudi_etal_2006, mcelreath_and_boyd_2007}. The rich cross-fertilisation between different subfields has promoted the designing of new analytical techniques as well as new uses for existing methods, but more crucially have prompted the development of theoretical models of how culture is acquired and transmitted between individuals~\citep{henrich_mcelreath2003,laland2004,mcelreath_and_boyd_2007,shennan_2009_patternprocess,Kempe_and_Mesoudi_2014,mesoudi_cultural_2015}.

@@

This paper examines a group of social learning strategies generally referred to as \emph{payoff-based transmission} \citep{schlag1998,kendal_etal_2009,lake_and_crema_2012,baldini2013,kandler_and_laland_2013,crema_lake_inpress}. Here the probability of copying a cultural variant $i$ is directly proportional to a payoff $s$ exhibited by individuals possessing $i$. In other words the social learner assumes that the success of an individual (measured in $s$) is a function (of which details are often unknown) of the observed behavioural trait, that is $s=f(i)$. Several variants of this learning strategies have been proposed and explored, including: \emph{imitate the best}, where the social learner identifies and copies the trait possessed by the individual with the highest payoff; \emph{compare means}, where the social learner compare the average payoff of different strategies and chooses the highest; and the \emph{proportional imitation rule}, where the probability of adopting a variant is proportional to the difference in payoff between the demonstrator and the learner~\citep{schlag1998,baldini2013,crema_lake_inpress}.

@@

The uncertainty in the payoff signal is derived by the presence of interacting cultural and physical traits, as well as the environment where the trait is expressed. Thus the function $s=f(i)$ should be better described as $s=f(i,\epsilon)$, where $\epsilon$ represent all additional hidden and latent variables that contribute, along with the cultural variant of interest $i$, to the payoff signal $s$. A trivial example is a fisherman who might evaluate the performance of a new hook (the target cultural trait) using as a cue the number of fish captured (the payoff signal). The correlation between the two is partly derived by the efficiency of the hook, but also by the rod, the choice of the bait, the amount of available prey, the choice of the fishing spot, the physical strength of the fisherman, and chance. In many cases we should expect that the social teacher and learner share many of these interacting traits as well as the environment where these are displayed. In other words, cultural and ecological inheritance should in part reduce the impact played by $\epsilon$, and consequently the uncertainty in the payoff-signal. There are however situations the payoff is highly variable even holding these variables constant. One of the most relevant examples are showcased in situations where the performance of a given  cultural variant $i$ is partly determined by the number of individuals $n_i$ expressing $i$ within a population. This \emph{density dependence}, which can be described by the payoff function $f(i,n_i)$, implies that the adoption (or abandonment) of a trait can alter its expected payoff-signal, potentially causing a cascade of feedback loops.

Several behavioural traits are expected to show such a pattern. For example, behaviour linked to the exploitation of limited resources will exhibit a decline in payoff when density exceeds a critical threshold (i.e carrying capacity). In this case new optimal variants are likely to spread rapidly when they are rare, but as the number of individuals possessing them approach their respective carrying capacities, their beneficial effects will start to decline, the payoff will decrease, and eventually rarer suboptimal variants might be preferred (see Lake and Crema \citeyear{lake_and_crema_2012} for an extensive analysis).

Density dependence does not necessarily imply only a negative correlation between population size and payoff. In many ecological contexts per capita growth rate is positively correlated with population density, thus resulting into an inverse density dependence generally linked with processes such as cooperation, mutual facilitation, and environmental conditioning. This phenomena, known as \emph{Allee} effect \citep{allee1958,courchamp_etal_1999}, has been extensively studied  from both empirical and theoretical standpoints, and its consequences recognised in a wide variety of contexts, including habitat selection \citep{greene_and_stamps_2001}, resource competition \citep{jang2013},  dispersal \citep{steele_2009}, shifts in settlement pattern \citep{crema_2014}, colonisation events  \citep{kennet_etal_2006}, genetic diversity \citep{roques_etal_2012}, and extinction of languages \citep{sutherland_2003}. Yet, despite its ubiquity, we are unaware of any theoretical studies exploring how the simultaneous presence of direct and inverse density dependence in the payoff signal can affect the evolutionary dynamics of cultural transmission.

Indeed, while the concept of Allee effect is well established in population ecology (see Kramer et al \citeyear{kramer_etal_2009} for a systematic review), there is no direct equivalent in the social sciences. Yet, there are multiple, established lines of evidence that support  the existence of both direct and inverse density dependence in human cultural and behavioural traits. For example several ethnographic studies on small scale societies show how the per-capita return of hunting groups show higher yield when the group size is not too small nor too big~ \citep{hill_and_hawkes_1983,janssen_and_hill_2014}. Other instances of inverse density dependence include population level activities such as the use of anthropogenic fire~\citep{bird2013} or selective hunting~\citep{dods_2002} which often require a population size above a certain critical density in order to promote sufficient niche enhancement and a consequent increase in payoff~\citep{rowley-conwy_and_layton_2011}. In a broader sense any cultural trait that benefits from direct or indirect cooperation, 
promote positive niche construction \citep{vandermeer_2008}, and requires a critical mass \citep{rogers_2003} are likely to exhibit some degree of Allee effect. The implication is thus not limited to traits that directly enhance reproductive fitness, and can include language~\citep{kandler2009}, organisational populations~\citep{caroll_and_hannan_1989} and communication technologies \citep{van_slyke_perceived_2007}.

Here we explore the broad consequences of the Allee effect by means of a theoretical equation-based model of cultural transmission. For the context of this study we narrow our scope by assuming that: 1) the social learning strategy is payoff-based; and 2)  the payoff is equivalent to the reproductive fitness of each individual. We demonstrate that direct and inverse density dependence can strongly affect the evolutionary trajectory of the system, leading to changes in the conditions required to reach specific equilibria, as well as to the emergence of unstable patterns characterised by temporary or permanent episodes of shifts between alternative cultural variants.

\section{The Model}
We consider the evolutionary dynamics of two mutually exclusive cultural variants which payoff functions are characterised by an Allee effect. We assume a non-fixed population of size $N$, subdivided into two subpopulations with sizes $a$ and $b$, each associated with a different cultural variant. We further assume that $a$ and $b$ can change as a result of intrinsic population growth/decline as well as shifts from one to the other as a result of cultural transmission. The core model can be depicted with the following coupled difference equation \eqref{eq1}:

@@

We investigate the equilibrium conditions for different initial population sizes of $a$ and $b$ and different settings of $z$, $A$, and $K$, ensuring that in all cases $K_b \geq K_a$. We assumed that two strategies differed only in their critical densities and carrying capacities but not on their intrinsic payoffs. To achieve this we set $r_b=\rho r_a$, where $\rho$ is a constant\footnote{Given two variants with the same baseline growth rate (i.e. $r_a=r_b$), equation \eqref{eq2} returns a lower $R$ for the variant with higher $A$ and $K$. In order to avoid this we increased the baseline of the strategy with lower $R$ by a factor $\rho$, given by the following equation:
\begin{align*}
\rho=\frac{A_bK_b(K_a-A_a)^2}{A_aK_a(A_b-K_b)^2}
\end{align*}
  } 
that ensures that the two variants have always the same $R_{max}$. Given that equations \ref{eq1}-\ref{eq4} have no analytical solution that we are aware of, we identified equilibrium conditions via simulation (source codes of the simulation can be found on the following repository \textcolor{red}{PUT LINK HERE})
%%% We Need to clean the code and make a github repository for this %%%

@@

\begin{figure*}
  \centering
      \includegraphics[width=\textwidth]{./figures/figure5}
  \caption{Basins of attraction for $z=0.05$ (top) and $z=0.2$ (bottom); $\lambda=1$ (left), $\lambda=0.75$ (middle) and $\lambda=0.5$ (right)}
    \label{fig:overlap}
\end{figure*}

@@

Our analysis examined the impact of direct and inverse density dependence in payoff-based social learning strategies. The results showed that Allee effect lead to the emergence of both positive and negative feedback mechanism, and can generate a wide range of patterns where small changes in the initial conditions can both deeply affect the evolutionary trajectories of the system. Our theoretical model indicated that social learning can in fact alter the basins of attraction of stable nodes, leading to different dynamics and equilibria from the very same initial conditions. If there is no reliance on social learning, the critical density separates the ultimate equilibrium of the system: if the initial condition is below this threshold the variant is destined to go extinct, whilst if it is above this value will eventually reach carrying capacity. When social learning is present this boundary shifts, variants can survive even if its initial condition is below critical density and equally can reach extinction even if above this threshold. This scenario is further enhanced when the two variants are not independent but are in direct competition. In this case, social learning can further enhance the advantage of one variant over the other.

When the reliance on social learning is comparatively high, and the two variants have a substantial overlap in their range of positive payoff densities (i.e. they share similar critical densities and carrying capacities), the system does not always reach an stable equilibrium, but can instead exhibit cyclical fluctuations between the two variants. In other words the impact of social learning is not absorbed by the system, and instead a negative feedback process leads to continuous switch from one variant to the other.

The implications of our model can provide some insights in the evolutionary history of cultural traits exhibiting density dependence. Perhaps the most notable example is the spread of keystone innovations in subsistence strategies which led to the adoption of variants enabling larger carrying capacity, often at the cost of higher critical densities. The spread of behavioural traits such as the know-how of the domestication process~\citep{barker2006} the use of new subsistence tools~\citep{petraglia_population_2009} and processing techniques~\citep{molleson1993}, as well as the inclusion/exclusion and extensification/intensification of different prey species are just some of the most notable examples. In many of these examples the initial spread of the innovation generated a positive feedback process, either through enhanced possibility of direct cooperation, or some form of mutual facilitation often indirectly emerging from positive niche construction. At the same time, ethnographic and archaeological evidence show how many instances of critical transitions where not smooth, often leading to the coexistence of alternative strategies as well as episodes of a return to a previously abandoned cultural traits (see Day and Walter \citeyear{day_and_walter_1989} for similar insights in economics). Several ethnographic examples showcase small-scale societies alternating between different subsistence modes over a comparatively brief period of time \citep{layton1991,mace_1993,oota_et_al_2005}, while increasing archaeological evidence suggest that the transition from hunting-gathering to farming was not unidirectional, but with instances of temporary reversions \citep{rowley2001,stevens_fuller_2012}. Formal cross-cultural analysis of these episodes are still rare (but see Ullah et al. \citeyear{ullah_etal_2015}) but the insights of our model suggests that different levels of reliance on social learning could alone portray some of the evolutionary trajectories observed in these systems.
